Reproducible development environments with rix

### 2024-10-01 paper.qmd
bootstrap env

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clearly list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

*Linked and executable code and data* refer to the compiled source code scripts,
but for an interpreted language such as [R]{}{.proglang}, this would rather
refer to the *computational environment*. The computational environment is all
the software required to run the analysis. Here too, we can speak of a
continuum. One could simply name and list the software used: for example, the
[R]{.proglang} programming language. Sometimes, authors have the courtesy to
also state the version of [R]{.proglang} used. Some authors go further, and also
list the packages used, and ideally with their versions as well. Authors rarely
state the operating system on which the analysis was done, even though it has
been shown that running the same analysis with the same software but on
different operating systems could lead to different results, as described in
@neupane2019. Authors also only very rarely provide instructions to install the
required tools and software in order to reproduce their studies.

@@

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.pkg} package manager for reproducibility, we developed [rix]{.pkg}.
[rix]{.pkg} comes with the `rix()` function that helps generate such expressions
quite easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

```{r, eval = F}
rix_init(
   project_path = path_subshell,
   rprofile_action = "overwrite",
   message_type = "simple"
 )
```

```{r, eval = F}
rix_init(
   project_path = path_subshell,
   rprofile_action = "overwrite",
   message_type = "simple"
 )
```

@@

We should say that `rix_init()` and `with_nix()` print many status messages to
the console while running, which we don't show here for space reasons.

The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
and is indeed different than the one generated from a recent version of R.

@@

There are many tools to make one’s analyses reproducible, but Nix distinguishes itself from 
these in that its main concern is in deploying
the complete set of packages that are *closed under the ''depends on''
relation*: Nix does not simply install one piece of software,
but its dependencies, and their dependencies. This makes Nix extremely
useful for reproducible research, but because it solves a complex problem,
it is a complex tool. With [rix]{.pkg}, we wish to make using Nix more
accessible by providing a familiar interface and 
user experience to [R]{.proglang} users.

>>>>

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clearly list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the [Nix]{.proglang} package manager, and an [R]{.proglang} package 
  called [rix]{.pkg} that makes using [Nix]{.proglang} much simpler 
  for users of [R]{.proglang}.

@@

*Linked and executable code and data* refer to the compiled source code scripts,
but for an interpreted language such as [R]{.proglang}, this would rather
refer to the *computational environment*. The computational environment is all
the software required to run the analysis. Here too, we can speak of a
continuum. One could simply name and list the software used: for example, the
[R]{.proglang} programming language. Sometimes, authors have the courtesy to
also state the version of [R]{.proglang} used. Some authors go further, and also
list the packages used, and ideally with their versions as well. Authors rarely
state the operating system on which the analysis was done, even though it has
been shown that running the same analysis with the same software but on
different operating systems could lead to different results, as described in
@neupane2019. Authors also only very rarely provide instructions to install the
required tools and software in order to reproduce their studies.

@@

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.pkg} package manager for reproducibility, we developed [rix]{.pkg}.
[rix]{.pkg} comes with the `rix()` function that helps generate such expressions
quite easily. [rix]{.pkg} is available on CRAN and can be installed as any other
package, but it is also possible to use it to bootstrap an [R]{.proglang}
development environment on a system where [R]{.proglang} is not available (but
[Nix]{.proglang} is). One simply needs to run this line inside a terminal:

```
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
```

to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
available. From there, it is possible to generate new [Nix]{.proglang}
expressions to build new environments. For example, here is how one could use
[rix]{.pkg} to generate a `default.nix` file that installs version 4.3.1 of
[R]{.proglang} and the [dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't
show the generated expression here because of space):

@@

@@

We should say that `with_nix()` prints many status messages to the console while
running, which we don't show here for space reasons.

The `expr` argument of `with_nix()` can be arbitrarily complex, but must be
wrapped inside a `function()` call. The value of `out_subshell` is `r
out_subshell`, and is indeed different than the one generated from a recent
version of R.

@@

There are many tools to make one’s analyses reproducible, but Nix distinguishes
itself from these in that its main concern is in deploying the complete set of
packages that are *closed under the ''depends on'' relation*: Nix does not
simply install one piece of software, but its dependencies, and their
dependencies. This makes Nix extremely useful for reproducible research, but
because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
wish to make using Nix more accessible by providing a familiar interface and
user experience to [R]{.proglang} users. Because it is possible to build
reproducible development shells using [Nix]{.proglang}, it can accomodate any
use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
and serve [plumber]{.pkg} application programming interfaces.
